---
title: Probabilistic forecasts for anomaly detection
author: Rob J Hyndman
date: 3 July 2024
titlegraphic: bg-06.png
toc: false
abstract: When a forecast is very inaccurate, it is sometimes because a poor forecasting model is used, but it can also occur when an unusual observation occurs. I will discuss the latter situation, where a good forecasting model can be used to identify anomalies. The approach taken is to use a probabilistic forecast, and to compute the "density scores" equal to the negative log likelihood of the observations based on the forecast distributions. The density scores provide a measure of how anomalous each observation is, given the forecast density. A large density score indicates that the observation is unlikely, and so is a potential anomaly. On the other hand, typical values will have low density scores. A Generalized Pareto Distribution is fitted to the largest density scores to estimate the probability of each observation being an anomaly. Applications to tourism numbers and mortality data will be used to illustrate the ideas using the weird R package.
format:
  presentation-beamer:
    template-partials:
        - before-title.tex
    knitr:
      opts_chunk:
        dev: "CairoPDF"
execute:
  cache: false
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| echo: false
#| include: false
library(knitr)
library(targets)
library(fpp3)
tar_load(c("pbs", "pbs_stretch", "pbs_fit", "pbs_fc", "pbs_scores"))
tar_load(c("fr_mortality", "fr_scores", "fr_anomalies", "fr_fit", "fr_augment"))
# Add graphics files to ensure dependency
tar_load(c("paradigm1","paradigm2","paradigm3","n01","as","ase"))

# Colours to be viridis for continuous scales and Okabe for discrete scales
options(
  digits = 3,
  width = 75,
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
# Fira Sans font for graphics to match the main text
ggplot2::theme_set(ggplot2::theme_get() + ggplot2::theme(text = ggplot2::element_text(family = "Fira Sans")))
```

## Time series anomaly detection paradigms

\begin{textblock}{7.2}(0.1,1.1)
\fontsize{13}{15}\sf
\begin{enumerate}\tightlist\itemsep=0.2cm
\item \only<4>{\textcolor{Orange}}{\textbf{Identify anomalies within a time series in real time}:\newline
  use one-step forecast distributions}
\only<2-4>{\item \textbf{Identify anomalies within a time series in historical data}:\newline
  use residual distributions from smoothing method}
\only<3-4>{\item \textbf{Identify an anomalous time series in a collection of time series}:\newline
  use feature-based approach}
\end{enumerate}
\end{textblock}
\placefig{7.5}{1.2}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm1.png}
\only<2-4>{\placefig{7.5}{3.9}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm2.png}}
\only<3-4>{\placefig{7.5}{6.5}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm3.png}}

## Australian PBS data

```{r}
#| label: pbs
pbs
```

## Australian PBS data

```{r}
#| label: pbs2
#| echo: false
tar_read(a12plot)
```

## Anomaly score distribution
\fontsize{12.5}{14}\sf

\only<1-3>{\begin{textblock}{10.2}(0.3,1.0)
\begin{block}{One-step forecast distribution: $N(\mu_t, \sigma^2)$}
\centerline{$\displaystyle f(y_t | y_1,\dots,y_{t-1}) = \phi\left(\frac{y_t - \mu_t}{\sigma}\right) = \frac{1}{\sigma\sqrt{2\pi}}\exp\left\{\frac{(y_t-\mu_t)^2}{2\sigma^2}\right\}$}
\end{block}
\end{textblock}
\placefig{10.8}{1.2}{width = 5.1cm}{figures/n01.png}
}

\only<2-3>{\begin{textblock}{10.2}(0.3,3.6)
\begin{block}{Anomaly score distribution: $S \sim \frac12 \chi^2_1 + c$}
\centerline{$\displaystyle s_t = -\log f(y_t | y_1,\dots,y_{t-1}) = \frac{1}{2}\left(\frac{y_t - \mu_t}{2\sigma}\right)^2 + \frac{1}{2}\log(2\pi\sigma^2)$}
\end{block}
\end{textblock}
\placefig{10.8}{3.8}{width = 5.1cm}{figures/as.png}
}

\only<3>{\begin{textblock}{10.2}(0.3,6.15)
\begin{alertblock}{Extreme anomaly score distribution}
\centerline{$H(x) = P(S \le u + x \mid S > u)$}
\begin{itemize}\tightlist
\item[\textcolor{black}{$\rightarrow$}] Generalized Pareto Distribution for almost all forecast distributions $f$.
\end{itemize}
\end{alertblock}
\end{textblock}
\placefig{10.8}{6.4}{width = 5.1cm}{figures/ase.png}
}

## Anomaly detection algorithm

For each $t$:

* Estimate one-step forecast density: $f(y_t | y_1,\dots,y_{t-1})$.
* Anomaly score: $s_{t} = -\log\hat{f}(y_{t}| y_1,\dots,y_{t-1})$.
* High anomaly score indicates potential anomaly.
* Fit a Generalized Pareto Distribution to the top 10% of anomaly scores seen so far.
* $y_t$ is anomaly if $P(S > s_t) < 0.05$ under GPD.

## Example

```{r}
#| label: a12
#| fig-height: 3.6
a12 <- pbs |> filter(ATC2 == "A12", Month <= yearmonth("2006 Jan"))
a12plus <- pbs |> filter(ATC2 == "A12", Month <= yearmonth("2006 Feb"))
fc <- a12 |> model(ets = ETS(Scripts)) |> forecast(h = 1)
```

\pause\vspace*{-0.315cm}

```{r}
#| label: a12plot0
#| echo: true
#| eval: false
fc |> autoplot(a12)
```

```{r}
#| label: a12plot1
#| echo: false
#| eval: true
#| fig-height: 3.6
tar_read(a12plot1)
```

\only<3>{\begin{textblock}{10}(2.1,3.6)\fontsize{12}{13}\sf
\begin{alertblock}{}
Forecast distribution:\hfill $N(70.5, 3.9^2)$
\end{alertblock}
\end{textblock}
}

## Example

```{r}
#| label: a12plus
#| eval: false
#| echo: true
a12 <- pbs |> filter(ATC2 == "A12", Month <= yearmonth("2006 Jan"))
a12plus <- pbs |> filter(ATC2 == "A12", Month <= yearmonth("2006 Feb"))
fc <- a12 |> model(ets = ETS(Scripts)) |> forecast(h = 1)
fc |> autoplot(a12plus)
```

```{r}
#| label: a12plot2
#| echo: false
#| eval: true
#| fig-height: 3.6
tar_read(a12plot2)
```

\begin{textblock}{10}(2.1,3.6)\fontsize{12}{13}\sf
\begin{alertblock}{}
Forecast distribution:\hfill $N(70.5, 3.9^2)$\only<2>{\\
Actual:\hfill $y_t = 36.4$\\
Anomaly score:\hfill $s_t = -\log f(y_t | y_1,\dots,y_{t-1}) = 40.8$ \\
Probability not an anomaly:\hfill $P(S > s_t) = 0.0154$}
\end{alertblock}
\end{textblock}

## Rolling origin forecasts

```{r}
#| label: tscvplots
#| echo: false
tar_read(tscvplot)
```

## Rolling origin forecasts

```{r}
#| label: pbsstretch
#| echo: true
#| eval: false
pbs_stretch <- stretch_tsibble(pbs, .step = 1, .init = 36)
```

```{r}
#| label: pbsstretch2
#| echo: false
pbs_stretch
```

## Rolling origin forecasts

```{r}
#| label: pbsfit
#| echo: true
#| eval: false
pbs_fit <- pbs_stretch |> model(ets = ETS(Scripts))
```

```{r}
#| label: pbsfit2
#| echo: false
pbs_fit
```

## Rolling origin forecasts

```{r}
#| label: pbsfc
#| echo: true
#| eval: false
pbs_fc <- forecast(pbs_fit, h = 1)
```

```{r}
#| label: pbsfc2
#| echo: false
#| eval: true
pbs_fc
```

## PBS anomalies

```{r}
#| label: pbs_scores
#| echo: true
#| eval: false
pbs_scores <- pbs_fc |>
  left_join(pbs |> rename(actual = Scripts), by = c("ATC2", "Month")) |>
  mutate(
    s = -log_likelihood(Scripts, actual),
    prob = lookout(density_scores = s, threshold = 0.9)
  )
```

```{r}
#| label: pbs_scores2
#| echo: false
#| eval: true
pbs_scores
```

## PBS anomalies 

```{r}
#| label: pbs_scores3
pbs_scores |> filter(prob < 0.05)
```

## PBS anomalies

```{r}
#| label: pbs_scores4
#| echo: false
tar_read(l03)
```

## PBS anomalies

```{r}
#| label: pbs_scores6
#| echo: false
tar_read(r06)
```


## PBS anomalies

```{r}
#| label: pbs_scores5
#| echo: false
tar_read(n07)
```

## Time series anomaly detection paradigms

\begin{textblock}{7.2}(0.1,1.1)
\fontsize{13}{15}\sf
\begin{enumerate}\tightlist\itemsep=0.2cm
\item {\textbf{Identify anomalies within a time series in real time}:\newline
  use one-step forecast distributions}
\item \only<2>{\textcolor{Orange}}{\textbf{Identify anomalies within a time series in historical data}:\newline
  use residual distributions from smoothing method}
\item \textbf{Identify an anomalous time series in a collection of time series}:\newline
  use feature-based approach
\end{enumerate}
\end{textblock}
\placefig{7.5}{1.2}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm1.png}
\placefig{7.5}{3.9}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm2.png}
\placefig{7.5}{6.5}{width = 8.6cm, height = 2.4cm, keepaspectratio = false}{figures/paradigm3.png}

## Example: French mortality

```{r}
#| label: fr_mortality
fr_mortality
```

## Example: French mortality

```{r}
#| label: fr_mortality_time_plots
#| echo: false
tar_read(fr_mortality_time_plots)
```

## Example: French mortality

```{r}
#| echo: true
#| eval: false
fr_fit <- fr_mortality |>
  model(stl = STL(log(Mortality)))
```

```{r}
#| label: fr_fit
fr_fit
```

## Example: French mortality

```{r}
#| echo: true
#| eval: false
augment(fr_fit)
```

```{r}
#| echo: false
#| eval: true
fr_augment
```

## Example: French mortality

```{r}
#| echo: true
#| eval: false
fr_sigma <- augment(fr_fit) |>
  group_by(Age, Sex) |>
  summarise(sigma = IQR(.innov)/1.349, .groups = "drop")
```

```{r}
#| echo: false
#| eval: true
tar_read(fr_sigma)
```


## Example: French mortality

```{r}
#| echo: true
#| eval: false
fr_scores <- augment(fr_fit) |>
  left_join(fr_sigma) |>
  mutate(
    s = -log(dnorm(.innov / sigma)),
    prob = lookout(density_scores = s, threshold_probability = 0.9)
  )
```

```{r}
#| echo: false
#| eval: true
fr_scores
```


## Example: French mortality

```{r}
#| label: fr_scores3
fr_scores |> arrange(prob)
```

## Example: French mortality

```{r}
#| label: fr_scores4
#| echo: false
tar_read(fr_anomalies_plot)
```

## Example: French mortality

```{r}
#| label: fr_scores5
#| echo: false
tar_read(fr_anomalies_plot2)
```

\def\frenchevents{\begin{textblock}{4.5}(-0.25,1.6)
\fontsize{12}{14}\sf
\begin{alertblock}{}
\begin{itemize}\parskip=0cm\itemsep=5pt
\item 1870--1872: Franco-Prussian war and repression of the ‘Commune de Paris’
\item 1914--1918: World War I
\item 1918: Spanish flu
\item 1939--1945: World War II
\end{itemize}
\end{alertblock}
\end{textblock}}
\only<2>{\frenchevents}

## Example: French mortality

```{r}
#| label: fr_scores6
#| echo: false
tar_read(fr_anomalies_plot3)
```

\frenchevents

## Example: French mortality

```{r}
#| label: fr_25
#| echo: false
tar_read(fr_25)
```

## More information

\placefig{1}{1.35}{width=3.5cm}{fable}
\placefig{4.5}{1.35}{width=3.5cm}{weird}

\only<2->{\begin{textblock}{8.8}(0.3,5.7)
\begin{alertblock}{}\fontsize{14}{17}\sf
\textbf{Slides}: \hfill\href{https://robjhyndman.com/isf2024}{robjhyndman.com/isf2024}

\textbf{Incomplete book}: \hfill\href{https://OTexts.com/weird}{OTexts.com/weird}

\textbf{fable}: \hfill\href{https://fable.tidyverts.org}{fable.tidyverts.org}

\textbf{weird}: \hfill\href{https://pkg.robjhyndman.com/weird}{pkg.robjhyndman.com/weird}

\end{alertblock}
\end{textblock}}

\only<3->{\begin{textblock}{6.1}(9.6,5.7)
\fontsize{12}{13}\sf
\begin{block}{Find me at ...}
\href{https://robjhyndman.com}{\faicon{home} robjhyndman.com}

\href{https://twitter.com/robjhyndman}{\faicon{twitter} @robjhyndman}

\href{https://github.com/robjhyndman}{\faicon{github}  @robjhyndman}

\href{mailto:rob.hyndman@monash.edu}{\faicon{envelope} rob.hyndman@monash.edu}
\end{block}
\end{textblock}}
